---
title: "Data Visualisation with ggplot2"
---

We're going to create some plots using functions from the `ggplot2` package. R packages usually contain functions related to a specific activity, which in this case is data visualisation, and also documentation which outlines how the functions work.

We've already installed the `ggplot2` package. It was installed as part of the `tidyverse` package, which is a popular collection of packages developed by the same group of people. 

# Reading In Data

Each time we start a new RStudio session, we need to load the packages that we want to use, and to do this we use `library()`.

```{r, warning=FALSE}
library(tidyverse)
```

After loading packages, the next thing we often need to do is to load some data into R. To load our spreadsheet in we use the `read_csv()` function from the `readr` package (which is loaded as part of the tidyverse package), and we'll store it in an object called `surveys`.`

```{r}
surveys <- read_csv(file = "data_raw/surveys_complete.csv")
```

You'll notice some useful information printed in the console, including the number of rows and columns, and the number of different column types. `chr` refers to character, or columns containing text values, while `dbl` refers to double, another name for numeric values.

Now we can see `surveys` in our environment pane and we can click on it to take a look.

This dataset is taken from the Portal Project, a long-term study from Portal, Arizona, in the Chihuahuan desert, which measures features of desert-dwelling mammals. Click on the `surveys` object in the environment pane to see the data.

It's worth pointing out, anything we do to the data in R is not changed in the file we read in. 

To get a bit more information about any object in R, we can look at it's structure:

```{r}
str(surveys)
```

# Plotting with ggplot2

The gg in ggplot2 stands for “grammar of graphics”, which refers to a particular way of creating plots using a consistent vocabulary and layering different elements on. Therefore, we only need small changes to our code if the underlying data changes or we decide to make a box plot instead of a scatter plot. This approach helps you create publication-quality plots with minimal adjusting and tweaking.

ggplot plots are built step by step by adding new layers, which allows us to customise and test as we go. We're going to use the following template to build our plots:

`ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) + <GEOM_FUNCTION>()`

We need to change everything inside `<>`.

First we need to specify the data we want to use. Next, we can see that the ggplot() call has a function nested within it, aes(), which relates to aesthetic mappings. This is where we specify which parts of the data should be mapped to different axes on our plot. Finally, we need to specify the geometry we want to use, or in other words, the plot type. The geometry function we use is added to the end with a `+` sign.

Let's go through step by step:

```{r}
ggplot(data = surveys)
```

We get a blank plot because we haven’t told `ggplot()` which variables we want to correspond to parts of the plot.

```{r}
ggplot(data = surveys, mapping = aes(x = weight, y = hindfoot_length))
```

Now we’ve got a plot with x and y axes corresponding to variables from surveys. However, we haven’t specified how we want the data to be displayed. We do this using `geom_` functions, which specify the type of geometry we want, such as points, lines, or bars. We can add a `geom_point()` layer to our plot by using the + sign. We indent onto a new line to make it easier to read, and we have to end the first line with the + sign.

```{r}
ggplot(data = surveys, mapping = aes(x = weight, y = hindfoot_length)) +
  geom_point()
```

You may notice a warning that missing values were removed. If a variable necessary to make the plot is missing from a given row of data (in this case, hindfoot_length or weight), that observation can't be plotted. `ggplot2` warns us that this has happened to ensure we know there are missing values.

Another common type of message is an error, which means R can't execute the command. This is commonly due to misspelling, or missing punctuation such as brackets or commas.

We can change the colour of the points by supplying a `colour` argument inside the geometry function.

```{r}
ggplot(data = surveys, mapping = aes(x = weight, y = hindfoot_length)) +
  geom_point(colour = "blue")
```

We can also change the transparency of points to get a better idea of where data is concentrated. Here we change transparency of points to 20%.

```{r}
ggplot(data = surveys, mapping = aes(x = weight, y = hindfoot_length)) +
  geom_point(colour = "blue", alpha = 0.2)
```

We may also want to modify the x axis to a log scale.

```{r}
ggplot(data = surveys, mapping = aes(x = weight, y = hindfoot_length)) +
  geom_point(colour = "blue", alpha = 0.2) +
  scale_x_log10()
```

## Boxplots

If we wanted to use a parametric statistical test like ANOVA, first we would need to check some assumptions. One of these assumptions is that the data are normally distributed. We can get a quick idea of this using a boxplot. Let's look at hindfoot_length grouped by plot_type.

```{r}
ggplot(data = surveys, mapping = aes(x = plot_type, y = hindfoot_length)) +
  geom_boxplot()
```

We can flip the plot around to make it easier to read the longer labels.

```{r}
ggplot(data = surveys, mapping = aes(x = plot_type, y = hindfoot_length)) +
  geom_boxplot() +
  coord_flip()
```

If we wanted to split our plot into separate plots to look at how the distribution of data differs by sex we can use faceting. Here we'll supply the column name we want to facet by within the `vars()` function, inside the `facet_wrap()` function. We can also specify how we want the plots to be arranged by supplying either `nrow` or `ncol`.

```{r}
ggplot(data = surveys, mapping = aes(x = plot_type, y = hindfoot_length)) +
  geom_boxplot() +
  coord_flip() +
  facet_wrap(vars(sex), ncol = 1)
```

## Exporting plots

Once we're happy with our plot, we can save it using the `ggsave()` function. 

We need to specify the following arguments:

- `filename`: the filename we're using to save our plot, including the file type of the image (e.g. ".jpg", ".png", ".tif". 
- `plot`: the name of the plot object we want to save.

```{r, eval=FALSE}
ggsave("figures/hindfoot-plottype_sex.jpg", plot = final_plot)
```

You can also specify one of the below optional arguments to set where the image will be saved:

- `path`: the path of the directory we want to save our plot in (if not already in our filename).
- `create.dir`: whether to create a new directory if the directory specified in `path` doesn't exist.

### File type

The first consideration is the file type of the image. `ggplot` supports both vector and raster formats. Vector formats store images as instructions which means they can be resized without any loss of quality, whereas raster formats store images as fixed arrays of pixels, meaning they cannot be resized without losing quality. The JPEG format we used above is an example of a raster format - the image resolution is fixed when it is saved, so it cannot be resized very much without looking bad.

Supported vector formats:

- .pdf (Portable Document Format, when unflattened)
- .svg (Scalable Vector Graphics)
- .eps (Encapsulated PostScript)

Supported raster formats:

- .png (Portable Network Graphics) 
- .jpg or .jpeg (Joint Photographic Experts Group) 
- .tiff or .tif (Tagged Image File Format)
- .bmp (Bitmap)

Sometimes a journal will ask for figures in a vector format so that they can size the image appropriately based on the typesetting of the document. Otherwise, most of the time you'll select a raster format and set the size of the image yourself. To do this you need to use the following optional arguments:

- `width` & `height`: plot size in units defined by the units argument (defaults to inches).
- `units`: one of the following - "in", "cm", "mm", "px".
- `dpi`: plot resolution in dots per inch.

### Clarifying DPI

Confusion around DPI is common. People often conflate DPI with other attributes of images relating to their size, or to the supposed 'quality' of an image.

As background, there are three properties that determine the size of an image:

- Resolution: the pixel dimension of a screen or an object on a screen, expressed as width x height (e.g. 1920x1080). 
- Print size: the physical dimensions of an image when printed, typically measured in inches for photos.
- DPI: Dots per inch, or the density of ink dots printed onto paper.

To obtain the print size of an image, the resolution is divided by the DPI set inside the image metadata. For example, if an image has a resoltuion of 900x900 pixels, and the DPI in the image is set to 300, then the image will be printed at 3 x 3 inches. The DPI setting inside the metdata of a digital image has absolutely no bearing on how that image is represented on a screen, only when it is printed or added into something like a Word doc which shows you how a document will look when printed.

It's common for journals to ask for figures based only on a particular DPI (often 300), but this by itself is meaningless. You could send them an image with a resolution of 150 x 150 pixels and a DPI set to 300, but when printed it would only be half an inch by half an inch. What they really mean is they want images with "a resolution large enough to display the image at the chosen size when printed."

If you'd like to determine the size or resolution yourself, then the value you provide for `width` & `height` will be multiplied by the value you give for `dpi` to give the resolution. For example, let's say we need our plot to be 3 inches wide on the page and 300dpi.

```{r, eval=FALSE}
ggsave("figures/hindfoot-plottype_sex.jpg", 
       plot = final_plot,
       width = 3,
       units = "in",
       dpi = 300)
```

Or let's say we want it to be 1500 pixels wide and 1200 pixels high.

```{r, eval=FALSE}
ggsave("figures/hindfoot-plottype_sex.jpg", plot = final_plot,
       width = 5,
       height = 4,
       units = "in",
       dpi = 300)
```


